Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-11-27 paper/main.tex
adapt abstract to Plos template

%###################################################################################################
\begin{abstract}
    \aclp{abm} have become popular tools in the study of complex cellular systems.
    However, despite their popularity, it is often difficult to systematically estimate their
    parameters and perform advanced analyses, e.g., using the profile-likelihood.
    In this work, we study the mechanical properties of rod-shaped bacteria and provide techniques
    for estimating their parameters and performing identifiability analysis on them.
    Our approach may serve as a blueprint for a modular framework which can be adapted to other
    research questions.
    We use time series of microscopic images as input data and incorporate advanced effects as the
    elasticity of the rod.
    The developed model serves as a generalized foundation to describe multiple species of
    rod-shaped bacteria such as \acl{ecoli} or \acl{bsubtilis}.
\end{abstract}

@@

>>>>

\end{flushleft}

@@

%###################################################################################################
\section*{Abstract}
Elongated bacteria such as \acl{ecoli} and \acl{bsubtilis} possess distinct mechanical
and physiological properties that shape their growth, division, and collective behaviors.
Despite the importance of these organisms, existing computational frameworks rarely capture their
flexible filamentous morphology, limiting the ability to study their mechanics in a systematic and
quantitatively grounded way.
Agent-based models are well suited for representing individual cells and their interactions, yet
their application to flexible rod-shaped bacteria has been hindered by the absence of appropriate
mechanical formulations and by the lack of practical methods for estimating model parameters from
experimental data.
In this work, we introduce a computational model that represents elongated bacteria as deformable
rods capable of growth, elastic bending, and mechanical interactions.
We develop a set of imaging-based and geometry-based comparison methods that extract cell shape and
position information from segmented microscopy data and quantitatively compare it with model
simulations at the single-cell level.
These methods enable the estimation of model parameters across multiple experimental scenarios of
increasing complexity, including cases that exhibit substantial cell-to-cell variability.
Our results demonstrate that parameter estimation for complex agent-based models of flexible
bacteria is feasible and can yield biologically interpretable constraints on cellular mechanics.
The modeling and analysis framework is modular, generalizable, and readily adaptable to a broad
range of rod-shaped species.
By integrating our mechanical model into an extensible computational platform, we provide tools that
support more rigorous model construction, systematic calibration, and mechanistic interpretation.
This work establishes a foundation for quantitatively studying the mechanical behavior of
filamentous bacteria and paves the way for future applications ranging from biofilm formation to
multicellular
patterning.

\section*{Author summary}

\linenumbers
